\documentclass[idxtotoc,hyperref,openany]{labbook} % 'openany' here removes the gap page between days, erase it to restore this gap; 'oneside' can also be added to remove the shift that odd pages have to the right for easier reading

\usepackage[ 
  backref=page,
  pdfpagelabels=true,
  plainpages=false,
  colorlinks=true,
  bookmarks=true,
  pdfview=FitB]{hyperref} % Required for the hyperlinks within the PDF
  
\usepackage{booktabs} % Required for the top and bottom rules in the table
\usepackage{float} % Required for specifying the exact location of a figure or table
\usepackage{graphicx} % Required for including images
\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\newcommand{\HRule}{\rule{\linewidth}{0.5mm}} % Command to make the lines in the title page
\setlength\parindent{0pt} % Removes all indentation from paragraphs

\begin{document}

\labday{Monday, September 10, 2012 }

\section{Dynamic programming and DPP}
For concreteness, we focus on Nussinov's DP algorithm. The ideas should work for general DPs.  Suppose we have a fixed DNA/RNA sequence $S$ of length $L$. Let $y$ denote a particular secondary folding structure of the sequence. We try to follow the notations of the Supplementary Methods of Taskar's SDPP paper. 

The factor graph has $T=L(L-1)$ number of variable nodes, indexed by $t=(i,j), i<j$. The possible values for the variable node $(i,j)$ is the set of energies possible for the subsequence $S(i,j)$. There are also $L(L-1)$ factor nodes indexed by $(i,j), i<j$. The factor node $F(i,j)$ is connected to the variable nodes $\{(i+1, j-1), (i+k,j), (k+1, j)\}, i<k<j $. For $(i,j) \neq (1, L)$, the factor $F(i,j) = 1$ if the substructure $y(i,j)$ has a consistent folding (no mismatched base-pairs, etc), and $F(i,j)=0$ if the substructure is not consistent. For the "last" factor, $F(1,L)=\mbox{energy of the structure y}$.

\labday{Saturday, September 22, 2012}

To fill in a bit more and correct some mistakes from above.

We have a factor graph with $T=\frac{L(L-1)}{2}$ variable nodes, indexed by $t=(i,j)$. Similarly, there are $\frac{L(L-1)}{2}$ factor nodes that we denote by $F(i,j)$. The factor node  $F(i,j)$ is connected to the variable nodes $\{(i+1, j-1), (k,j), (i,k)\}, i<k<j $.  Generic variable nodes are denoted by $t$, and generic factor notes are denoted by $\alpha$. 

\begin{enumerate}
\item The values for $y_t$ are the possible folding scores for the subsequence $S(i,j)$. In the simplest case, the folding score equals to the number of base pairings, and $y_t \in \{0, 1, ..., L/2\}$.
\item For a factor node $\alpha$, we think of $y_{\alpha}$ as a list of values $\{y_{t_1}, y_{t_2}, ...\}$ where $t_i$ is a variable node connected to $\alpha$. Let $\alpha*$ denote the final factor node $F(1,L)$.
\item A factor node is associated with weight $w_{\alpha}(y_{\alpha})$. We have 
\[
w_{\alpha}(y_{\alpha})=(q^2(y_{\alpha}), q^2(y_{\alpha})\phi_r(y_{\alpha}), q^2(y_{\alpha})\phi_l(y_{\alpha}), q^2(y_{\alpha})\phi_r(y_{\alpha})\phi_l(y_{\alpha}) )
\]
\item For $\alpha \neq \alpha*$, $q(y_{\alpha})=1$ if the list $y_{\alpha}$ is feasible, i.e. the score $y_{ij}$ can be feasibly arrived from $y_{(i+1)(j-1)}, y_{kj}, y_{ik}$ and the sequence $S(i,j)$. And $q(y_{\alpha})=0$ otherwise. For $\alpha*$, $y_{\alpha*}=y_{1L}$, the score of the entire structure. 
\end{enumerate}

Questions to address:
\begin{enumerate}
\item The two pass belief propagation computes $\sum_{y} \prod_{\alpha} w_{\alpha}(y_{\alpha})$, where $y=\{y_{ij}\}$ is a list of scores over all substructures. What we actually need to compute is $\sum_{s} \prod_{\alpha} w_{\alpha}(y_{s})$, where the sum is over all feasible substructures. Seems like there is an straightforward bijection between $y$ and the set of all possible structures for S. Need to verify.
\item For the two pass belief propagation to work, i.e. equation 3 of the supplement, we need the factor graph to be a tree. This is not true in our case, which has many loops. However, the weights at the factor nodes are simple, $w_{\alpha}(y_{\alpha}=(0, 0, 0, 0))$ if $y_{\alpha}$ is not feasible. So might still work.
\item As we discussed, the factor graph is densely connected. But for a given structure, only a few variable nodes $y_t$ contribute. Need to work this out precisely.
\end{enumerate}

\labday{Tuesday, January 29, 2013}

More fleshing out of the previous notes.  Notation for variables and factors:
\begin{enumerate}
\item $y_{ij}$ is the value associated with the dynamic programming
  table entry $i,j$.
\item $d_{ij}$ is the decision variable for table entry $i,j$.  The
  decision variable tells what parent node was used to reach $i,j$.
  The value of $d_{ij}$ is just some $i',j'$.
\item $x_{ij} = \{0,1\}$ tells us whether entry $i,j$ is part of the
  solution path.
\item $\alpha_{ij}$ is the set of parents for table entry $i,j$.  This
  defines the factors.  We write $y_{\alpha}$ for the list of values
  $\{y_{t_1}, y_{t_2}, ...\}$ where $t_k$ is a variable node connected
  to $\alpha$. Let $\alpha*$ denote the final factor node.
\end{enumerate}

To apply SDPP-type stuff, we need to define a quality and diversity
score for each factor. For $\alpha \neq \alpha*$, the (non-negative)
quality score $q(y_{\alpha})=1$ if the list $y_{\alpha}$ is correctly
computed, that is, the score $y_{ij}$ follows from the decision
$d_{ij}$ and corresponding parent value $y_{i'j'}$ (and
$q(y_{\alpha})=0$ otherwise). For $\alpha*$, $q(y_{\alpha*})=y_{ij}$,
the score of the entire structure.  For any factor $\alpha$, the
(unit-length) diversity feature vector is simply an indicator
$\phi(y_{\alpha}) = x_{ij}$, that is, whether the entry $i,j$ is on
the solution path.  The quality and diversity functions for each node
are combined into a four-dimensional weight vector
$w_{\alpha}(y_{\alpha})$:
\[
w_{\alpha}(y_{\alpha})=(q^2(y_{\alpha}), q^2(y_{\alpha})\phi_r(y_{\alpha}), q^2(y_{\alpha})\phi_l(y_{\alpha}), q^2(y_{\alpha})\phi_r(y_{\alpha})\phi_l(y_{\alpha}) )
\]

Inference for the SDPP involves message-passing on the semi-ring with
operations $\oplus$ and $odot$.  We have messages from variables to
factors $m_{t \rightarrow \alpha}$ and from factors to variables
$m_{\alpha \rightarrow t}$.  The messages from variables to factors
are given by
\[
m_{t \rightarrow \alpha} = \otimes_{\alpha' \in F(t)\\ \alpha} m_{\alpha' \rightarrow t}(y_t)
\]
which isn't too bad to compute since each variable $y_t$ participates
in relatively few factors: decision-variables participate in only one
factor; $y_{ij}$ participate once as children and a few times as
parents (depending on the structure of the dynamic program).

and the messages from factors to variables are given by 
\[
m_{\alpha \rightarrow t}(y_t) = \oplus_{y_\alpha \sim y_t} [ w_\alpha( y_\alpha ) \otimes_{t' \in \alpha \\t} m_{t' \rightarrow \alpha} ]
\]
where we write $y_\alpha \sim y_t$ to indicate all sets $y_\alpha$
that are consistent with the given value of $y_t$.  At first, it may
seem complicated to compute the second set of messages because we must
consider all (exponentially many) sets $y_\alpha$ with a given setting
for $y_t$.  However, we can use some tricks because only a few
variables contribute to the quality and diversity scores.
Specifically, there are four cases:
\begin{enumerate}
\item $y_t = y_{ij}$ (child node).  For a given setting of the child
  $y_{ij}$, we need to consider all combinations of the path variable,
  the decision variable, and the corresponding parent.  The values of
  all the other possible parents don't contribute, so we can just
  multiply the inner term $[ w_\alpha( y_\alpha ) \otimes_{t' \in
      \alpha \\t} m_{t' \rightarrow \alpha} ]$ by the product of the
  number of settings for all the other parent variables.
\item $y_t = y_{i'j'}$ (a parent).  For each value of the parent, we
  consider two terms: first, when the parent is the corresponding
  parent given the setting of $d_{ij}$ (need to sum over all settings
  of $y_{ij}$) and all other settings (which is just a large product
  over the values that the child and other parents may take on).
\item $y_t = d_{ij}$.  We only need to consider all values of the
  child node and all values of the corresponding parent node,
  multiplied by the product of the number of settings for al the other
  parent variables.
\item $y_t = x_{ij}$. WE NEED TO ARTICULATE THIS ONE BETTER: I THINK
  WE NEED ANOTHER SET OF FACTORS CONNECTING $d_{ij}'s$ and $x_{ij}'s$
  that ensure consistency.
\end{enumerate}

If the factor graph has loops, then we must normalize the messages to
keep them from blowing up.  To do so, we
FILL IN HERE RE NORMALIZATION

Once we do the message-passing on the semi-ring, we can use the values
FILL IN HERE RE WHY WE DID THE MESSAGE PASSING
to compute the elements of the matrix $C$.
FILL IN HERE RE THE SDPP SAMPLER


\end{document}


